A thermodynamic view of dusty protoplanetary disks

### 2017-08-02 model.tex
revising according to AY's comments, up to 3.4

\section{Single fluid description of dusty gas}\label{setup} 
We model an accretion disk as a mixture of gas with dust treated as a 
pressureless fluid. We denote their density and
velocity field as $(\rhog,\bm{v}_\mathrm{g})$ and
$(\rhod,\bm{v}_\mathrm{d})$, respectively. The mixture has a single 
temperature $T$, and its pressure $P$ arise solely from the 
gas component. %neglect turbulence, viscosity, self-g and b fields.

The two fluids interact via a drag force parameterized by the relative stopping 
time $\tstop$ such that 
\begin{align}  
  \rhod\left.\frac{\p \bm{v}_\mathrm{d}}{\p t}\right|_\mathrm{drag}= -
  \rhog\left.\frac{\p \bm{v}_\mathrm{g}}{\p t}\right|_\mathrm{drag}=
  - \frac{\rhog\rhod}{\rho}\frac{\widetilde{\bm{v}}}{\tstop}. 
\end{align}
is the dust-gas friction force per unit volume, where
\begin{align}
  \widetilde{\bm{v}} \equiv\bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} 
\end{align}
is the dust-gas velocity difference; and 
\begin{align}
  \rho \equiv \rhog + \rhod
\end{align}
is the total density. 
Note that $\tstop$ 
differs slightly from the \emph{particle} stopping time $\tau_\mathrm{s} =
\tstop\rho/\rhog$ used in some studies
\citep[e.g.][]{youdin05a,jacquet11}.

In general $\widetilde{\bm{v}}$ obeys a complicated evolutionary equation
\citep[see, e.g.][]{youdin05a}. However, for tightly-coupled dust
particles with $\tstop\OmK\ll 1 $, where $\OmK$ is the Keplerian orbital
frequency, we can use the 
`terminal velocity approximation' to set %The dust-gas velocity difference 
%Then so that the dust-gas velocity
%difference is 
\begin{align}
  \widetilde{\bm{v}} = \frac{\nabla
    P}{\rhog}\tstop 
\end{align}
\citep{jacquet11}. This equation reflect the physical result that dust
particles tend to drift towards pressure maxima.

In the limit of strong dust-gas drag, $\tstop\to0$, the mixture obeys 
the first-order (in $\tstop$) single-fluid equations: 
\begin{align} 
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi_\mathrm{tot} - \frac{1}{\rho}\nabla  P, \label{momeq}\\ 
  &\frac{D\tepsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\tepsilon 
  \tstop \nabla P \right),\label{dusteq}\\
  &\frac{D T}{D t} = - \left(\gamma-1\right)T\nabla\cdot\bm{v} +
  \mathcal{H}_\mathrm{drag} + \mathcal{H}_\mathrm{eff}  - \Lambda  \label{tempeq} 
\end{align}
\citep{laibe14,price15}, where $D/Dt \equiv \p_t + \bm{v}\cdot\nabla$ is the Lagragian
derivative \emph{following the mixture} with center-of-mass velocity  
\begin{align}
  \bm{v} \equiv \frac{\rhog\bm{v}_\mathrm{g} + 
    \rhod\bm{v}_\mathrm{d}}{\rho}. 
\end{align}
For an ideal gas the pressure is given
by $P = \mathcal{R}\rhog T/\mu $, where $\mathcal{R}$ is
the gas constant and $\mu$ is the mean molecular weight.

Eq. \ref{dusteq} results from mass conservation of dust, where 
\begin{align}
  \tepsilon \equiv \frac{\rhod}{\rho}  = \frac{\epsilon}{1+\epsilon} 
\end{align}
is the dust-fraction and $\epsilon=\rhod/\rhog$ is the usual
dust-to-gas ratio. Note that Eq. \ref{dusteq} is 
equivalent to Eq. 46 of \cite{jacquet11}.

For the mixture's temperature evolution, Eq. \ref{tempeq}, $\gamma$ is
the adiabatic index; $\mathcal{H}_\mathrm{drag}$ represents heating
due to dust-gas drag and $\mathcal{H}_\mathrm{eff}$ is an effective
energy source arising from transforming from the two-fluid to
one-fluid description \citep[see][ for details]{laibe14}. We also
include a simple model of radiative cooling,
\begin{align}
  \Lambda = \frac{T -
    T_\mathrm{ref}}{t_\mathrm{cool}}, \label{cooling} 
\end{align}
which relaxes the mixture back to a prescribed temperature profile   
$T_\mathrm{ref}$ on a timescale of $\tcool$. We will shortly simplify
the problem by considering rapid cooling, $\tcool\to0$.

The total gravitational potential is $\Phi_\mathrm{tot}=\Phi +
\psi$. Here $\Phi$ originates from a star of mass $M_*$, 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right),  
\end{align}
where $(r,\phi, z)$ are cylindrical co-ordinates centered on the star, and 
$G$ is the gravitational constant. The second equality is the  
thin-disk approximation, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain explicit expressions for disk equilibria (\S\ref{eqm}).

The disk's own gravitational potential $\psi$ is included for
completeness, but we will mostly neglect it unless stated
otherwise. In such cases $\psi$ satisfies the Poisson equation 
\begin{align}
  \nabla^2\psi = 4 \pi G \rho.\label{poisson}
\end{align}

\subsection{Locally isothermal equation of state}\label{loc_iso_eos}
For later comparison it is convenient to re-write the temperature
evolution, Eq. \ref{tempeq}, to that for the pressure: 
\begin{align}
  \frac{\p P}{\p t} + \bm{v}_\mathrm{g}\cdot\nabla P = - \gamma P
  \nabla\cdot \bm{v}_\mathrm{g}  + \mathcal{H}\label{real_pressure}
%\frac{\rho \mathcal{R}}{\mu}\frac{T
%  - T_\mathrm{ref}}{t_\mathrm{cool}}, \label{realenergy}
\end{align}
where $\mathcal{H}$ collects together dust-drag heating and radiative
cooling.

We now simplify the gas energy equation by considering short cooling 
times $\tcool\to 0$, appropriate for the outer parts of an irradiated
protoplanetary disk \citep{chiang97,lin15}. Then the disk temperature
$T = T_\mathrm{ref}$ at all times, and so we may
adopt a locally isothermal equation of state 
\begin{align}\label{eos}
  P = c_s^2(r,z)\rhog = c_s^{2}\left(1 - \tepsilon\right)\rho,   
\end{align}
where $c_s(r,z)= \sqrt{\mathcal{R}T_\mathrm{ref}/\mu}$ is a prescribed
sound-speed profile fixed in time. In practice we consider vertically 
isothermal disks with \begin{align}\label{power_temp}
  c_s^2(r) \propto r^{q},
\end{align}
where $q$ is the power-law index for the disk temperature. For $q=0$
the disk is strictly isothermal.

Notice Eq. \ref{eos} resembles an ideal gas equation of state but with  
a reduced temperature $\widetilde{T} = 
T_\mathrm{ref}\left(1-\tepsilon\right)$. %sound-speed $c_{s,\mathrm{eff}} = c_s\sqrt{\left(1 -
This reduced temperature decreases with dust-loading. 
%   \tepsilon\right)}$. 
%then the equation of state has the standard
%form $P=c_{s,\mathrm{eff}}^2\rho$. 
Since $\tepsilon$ typically decrease away from the midplane, we expect
vertically isothermal dusty disks to behave as if it had a temperture
\emph{increasing} away from $z=0$.

\subsection{Effective energy equation}\label{energy_analogy}
Although we have eliminated the true energy
equation (Eq. \ref{tempeq}) by fixing a locally isothermal
equation of state (Eq. \ref{eos}), we show that the mixture 
nevertheless obeys an effective energy equation. This is
because advection of the dust-fraction, described by
Eq. \ref{dusteq}, can be transformed into an energy-like 
equation. 
%We show that for a prescribed temperature distribution, the mixture
%obeys an evolutionary energy equation, due to the advection of the
%dust-fraction. 
Eliminating $\tepsilon$ in favor of $P$, 
\begin{align*}
  \tepsilon = 1 - \frac{P}{c_s^2(r,z)\rho}, 
\end{align*}
then Eq. \ref{dusteq} becomes
\begin{align}
%\frac{DP}{Dt} 
\frac{\p P}{\p t} + \bm{v}\cdot\nabla P  
&= - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2}
                + \mathcal{C},  \label{eff_energy} \\
\mathcal{C}&\equiv c_s^2 \nabla\cdot\left[\tstop\left(1 -
  \frac{P}{c_s^2\rho}\right)\nabla 
  P\right].
\end{align}
Comparing with Eq. \ref{real_pressure}, we see that 
Eq. \ref{eff_energy} is the energy equation for an ideal gas of
adiabatic index $\gamma=1$, but now with the imposed temperature
gradient $\nabla c_s^2$ and dust-gas drag $\mathcal{C}$ playing the
role of radiative cooling ($\Lambda$ in 
Eq. \ref{tempeq}).

%In this form, the function $\mathcal{C}$ can be interpreted
%as a cooling term.    
If we denote 
\begin{align}
  \bm{F} \equiv  - \frac{\nabla P}{\rho}
\end{align}
for the pressure forces, then
\begin{align*}
  \mathcal{C} = - c_s^2\nabla \left( \tepsilon \tstop \rho \bm{F}
  \right), 
\end{align*}
which is in the same form as cooling by radiative diffusion. In protoplanetary
disks the corresponding heat flux, proportional to $\bm{F}$, is
directly radially outwards and vertically upwards. This is simply a
reflection of particle drift towards increasing pressure (inwards and 
downwards). Particle flux into a region contributes to `cooling' of
that region because the reduced temperature is lowered.

@@

Since we have shown that a locally isothermal dusty gas effectively
has $\gamma=1$ we can define an effective (dimensionless)    
entropy for the \emph{mixture} as 
\begin{align}
   \seff \equiv \ln \frac{P}{\rho} = \ln{\left[c_s^2(1-\tepsilon)\right]}.  
\end{align} 
%This follows naturally by combining the effective 
%energy Eq. \ref{eff_energy} and Eq. \ref{masseq}. 
Then 
\begin{align*}
  \frac{D \seff}{D t} = \bm{v}\cdot\nabla\ln{c_s^2} +
  \frac{c_s^2}{P}\nabla\cdot\left(\tstop\tepsilon\nabla P\right), 
\end{align*}
which is just the statement of entropy evolution in standard  
hydrodynamics, albeit with source terms. With an effective
entropy defined this way, many of the  
results concerning the stability of (locally) isothermal dusty gas
will have identical form and interpretations to that for pure gas.

@@

%\subsection{Dusty buoyancy forces}
Having identified the entropy of an isothermal dusty gas, we can also
define the disk vertical buoyancy frequency $N_z$ as   
%$N_z$ of the mixture is given by 
\begin{align}
  N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p \seff}{\p z}
  =c_s^2(r) \frac{\p \ln{\rhog}}{\p z}\frac{\p \tepsilon}{\p z},
%  \quad 
 % N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p \seff}{\p z}.
%&=
  %\frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\frac{\p\ln\rhog}{\p 
%  z}\frac{\p\epsilon}{\p z} \\ &
  %                                =
 % \frac{\epsilon}{\left(1+\epsilon\right)^2}\left(\frac{z}{\Htilde}\right)^2\OmK^2\notag,  
\end{align}
%Typically in a thin, smooth disk we have $|N_r|\ll |N_z|$. 
where the second equality assumes a vertically isothermal disk. A
similar definition exists for the radial buoyancy frequency
$N_r$. However, in thin, smooth disks we have $|N_r|\ll |N_z|$.

@@

%{\bf
\subsection{Physical disk conditions}
%physical disk conditions underwhich the above approx
%(small/well-coupled dust and isothermal gas) is applicable. smaller
%than mm at few tens of au. give dimensional numbers here
%}
The above correspondence between isothermal dusty gas and adiabatic
pure gas is derived under the terminal velocity approximation, which applies
to short stopping times, $\tstop\OmK\ll 1$; and the locally isothermal
approximation, which applies to short cooling times, $t_\mathrm{cool}\OmK\ll
1$. In typical protoplanetary disk
models such as the Minimum Mass Solar Nebula, $\tstop$ decreases for
smaller particles and/or at smaller radii \citep{youdin11}. However,
$t_\mathrm{cool}$ is generally only small in the outer disk
\citep{lin15,malygin17}. Combining these results, we estimate the 
correspondence applicable to particles less than mm in size at a few
to 10s of AU in protoplanetary disks. However, note that the
isothermal approximation may be relaxed to other fixed equations of
state (see \S\ref{gen_poly}).

@@

On the other hand, Eq. \ref{dusty_solberg2} can be violated in
disks if the dust is vertically well-mixed ($\p_z\epsilon =
0$) but  radially-varying ($\p_r\epsilon\neq0$). In this case the
left-hand-side of Eq. \ref{dusty_solberg2} becomes
\begin{align}
-\frac{c_s^4}{(1+\tepsilon)^4}\left(\frac{\p\ln{\rhog}}{\p
    z}\right)^2\left(\frac{\p\epsilon}{\p r}\right)^2<0.
\end{align}
Such an isothermal dusty disk is \emph{unstable} but there is no
effective vertical buoyancy to stabilize vertical motions ($N_z=0$)
that can tap into the free energy associated with  
vertical shear due to the radial gradient in the dust-to-gas ratio. 
This situation is identical to the pure gas, adiabtic simulations of \cite{nelson13}
where the gas entropy is vertically uniform but has a radial
gradient. The authors indeed find instability. We give a numerical
example in the dusty context in \S\ref{vert_mixed}.

>>>>

\section{Single fluid description of dusty gas}\label{setup} 
We model an accretion disk as a mixture of gas with dust treated as a 
pressureless fluid. We denote their density and
velocity field as $(\rhog,\bm{v}_\mathrm{g})$ and
$(\rhod,\bm{v}_\mathrm{d})$, respectively. The 
mixture has total density \begin{align}
  \rho \equiv \rhog + \rhod,
\end{align}
and center of mass velocity, 
\begin{align}
  \bm{v} \equiv \frac{\rhog\bm{v}_\mathrm{g} + 
    \rhod\bm{v}_\mathrm{d}}{\rho}, 
\end{align}
a single temperature $T$, and its pressure $P$ arise solely from the  
gas component. Our goal is to obtain a set of equations
describing the dust-gas mixture that resembles standard, single-phase
hydrodynamics. 
%This is possible by considering
%strongly coupled dust particles (but not neccessarily perfectly
%coupled) and an isothermal gas equation of state.

%\subsection{Terminal velocity approximation}

The dust and gas fluids interact via a drag force parameterized by the relative stopping 
time $\tstop$ such that 
\begin{align}  
  \rhod\left.\frac{\p \bm{v}_\mathrm{d}}{\p t}\right|_\mathrm{drag}= -
  \rhog\left.\frac{\p \bm{v}_\mathrm{g}}{\p t}\right|_\mathrm{drag}=
  - \frac{\rhog\rhod}{\rho}\frac{\left(\bm{v}_\mathrm{d} - \bm{v}_\mathrm{g}\right)}{\tstop}. 
\end{align}
is the dust-gas friction force per unit volume.%, where
%\begin{align}
%  \widetilde{\bm{v}} \equiv\bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} 
%\end{align}
%is the dust-gas velocity difference. 
 Note that $\tstop$ 
differs slightly from the \emph{particle} stopping time $\tau_\mathrm{s} =
\tstop\rho/\rhog$ used in some studies
\citep[e.g.][]{youdin05a}.

In general the relative velocity $\bm{v}_\mathrm{d} -
\bm{v}_\mathrm{g}$ obeys a complicated evolutionary equation 
\citep[see, e.g.][]{youdin05a}.% obtained from taking the difference
%between evolutionary equations for $\bm{v}_\mathrm{g}$ and
%$\bm{v}_\mathrm{d}$. 
 However, for tightly-coupled dust
particles with $\tstop\OmK\ll 1 $, where $\OmK$ is the Keplerian orbital
frequency, we can use the 
`terminal velocity approximation' to set %The dust-gas velocity difference 
%Then so that the dust-gas velocity
%difference is 
\begin{align}
  \bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} = \frac{\nabla
    P}{\rhog}\tstop 
\end{align}
\citep{jacquet11}. This equation reflects the well-known effect of 
particle drift towards pressure maxima \citep{weidenschilling77}.

%\subsection{One-fluid equations for dusty gas}

Under the terminal velocity approximation the dust-gas mixture obeys   
the first-order (in $\tstop$) one-fluid equations: 
\begin{align} 
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
   &\frac{D\tepsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\tepsilon 
     \tstop \nabla P \right),\label{dusteq}\\
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi_\mathrm{tot} - \frac{1}{\rho}\nabla  P, \label{momeq}\\ 
  &\frac{D T}{D t} = - \left(\gamma-1\right)T\nabla\cdot\bm{v} +
  \mathcal{H} + \mathcal{H}_\mathrm{eff}  - \Lambda  \label{tempeq} 
\end{align} 
\citep[see ][for a detailed derivation from the two-fluid
equations]{laibe14,price15}, where $D/Dt \equiv \p_t +
\bm{v}\cdot\nabla$ is the Lagragian 
derivative \emph{following the mixture}'s center-of-mass velocity $\bm{v}$. 
For an ideal gas the pressure is given 
by $P = \mathcal{R}\rhog T/\mu $, where $\mathcal{R}$ is
the gas constant and $\mu$ is the mean molecular weight.

Eq. \ref{dusteq} results from mass conservation of dust, where 
\begin{align}
  \tepsilon \equiv \frac{\rhod}{\rho}  = \frac{\epsilon}{1+\epsilon} 
\end{align}
is the dust-fraction and $\epsilon=\rhod/\rhog$ is the usual
dust-to-gas ratio. If $\tstop=0$ then the 
dust-to-gas ratio is conserved following the fluid. Otherwise, 
$\epsilon$ evolves due to particle drift in response to pressure
gradients. Note that Eq. \ref{dusteq} is equivalent to Eq. 46 of 
\cite{jacquet11}.

The total gravitational potential is $\Phi_\mathrm{tot}=\Phi +
\psi$. Here $\Phi$ originates from a star of mass $M_*$, 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right),  
\end{align}
where $(r,\phi, z)$ are cylindrical co-ordinates centered on the star, and 
$G$ is the gravitational constant. The second equality is the  
thin-disk approximation, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain explicit expressions for disk equilibria (\S\ref{steady_state}). 
The disk's own gravitational potential $\psi$ is included for
completeness, but we will mostly neglect it unless stated
otherwise. In such cases $\psi$ satisfies the Poisson equation 
\begin{align}
  \nabla^2\psi = 4 \pi G \rho.\label{poisson}
\end{align}

For the mixture's temperature evolution, Eq. \ref{tempeq}, $\gamma$ is
the adiabatic index; $\mathcal{H}$ represents heating (which would
include that due to dust-gas friction) and $\mathcal{H}_\mathrm{eff}$
is an effective source term arising from transforming from the
two-fluid to one-fluid formalism \citep[see][ for
details]{laibe14}. We also include a simple model of radiative
cooling, 
\begin{align}
  \Lambda = \frac{T -
    T_\mathrm{ref}}{t_\mathrm{cool}}, \label{cooling} 
\end{align}
which relaxes the mixture back to a prescribed temperature profile   
$T_\mathrm{ref}$ on a timescale of $\tcool$. We will shortly simplify
the problem by considering rapid cooling, $\tcool\to0$.

Eq. \ref{masseq}---\ref{tempeq} are not yet equivalent to standard 
hydrodynamics, which typically evolves two scalar fields: the density and
temperature (or pressure). By contrast, Eq. \ref{masseq}---\ref{tempeq}
involves 3 scalars: $\rho$, $\tepsilon$, and $T$. 
However, we can establish an  analogy by fixing the gas
equation of state, thus eliminating Eq. \ref{tempeq}; but then reform 
the dust-to-gas ratio evolution equation (\ref{dusteq}) into an
energy-like equation.

\subsection{Locally isothermal equation of state}\label{loc_iso_eos}
We simplify the gas energy equation by considering short cooling 
times $\tcool\to 0$, appropriate for the outer parts of an irradiated
protoplanetary disk \citep{chiang97,lin15}. Then the disk temperature
$T = T_\mathrm{ref}$ at all times, and so we may
adopt a locally isothermal equation of state 
\begin{align}\label{eos}
  P = c_s^2(r,z)\rhog = c_s^{2}\left(1 - \tepsilon\right)\rho,   
\end{align}
where $c_s(r,z)= \sqrt{\mathcal{R}T_\mathrm{ref}/\mu}$ is a prescribed
sound-speed profile fixed in time. In most applications we consider vertically 
isothermal disks with \begin{align}\label{power_temp}
  c_s^2(r) \propto r^{q},
\end{align}
where $q$ is the power-law index for the disk temperature. For $q=0$
the disk is strictly isothermal.

Notice Eq. \ref{eos} resembles an ideal gas equation of state but with  
a reduced temperature $\widetilde{T} = 
T_\mathrm{ref}\left(1-\tepsilon\right)$. %sound-speed $c_{s,\mathrm{eff}} = c_s\sqrt{\left(1 -
This reduced temperature decreases with dust-loading. 
%   \tepsilon\right)}$. 
%then the equation of state has the standard
%form $P=c_{s,\mathrm{eff}}^2\rho$. 
Since $\tepsilon$ typically decrease away from the midplane, we expect
vertically isothermal dusty disks to behave as if the temperture
\emph{increased} away from $z=0$.

\subsection{Effective energy equation}\label{energy_analogy}
Although we have deleted the true energy
equation (Eq. \ref{tempeq}) by fixing a locally isothermal
equation of state (Eq. \ref{eos}), we show that the mixture 
nevertheless obeys an effective energy equation. This is
because advection of the dust-fraction, described by
Eq. \ref{dusteq}, can be transformed into an energy-like 
equation. 
%We show that for a prescribed temperature distribution, the mixture
%obeys an evolutionary energy equation, due to the advection of the
%dust-fraction. 
Eliminating $\tepsilon$ in favor of $P$, 
\begin{align*}
  \tepsilon = 1 - \frac{P}{c_s^2(r,z)\rho}, 
\end{align*}
then Eq. \ref{dusteq} becomes
\begin{align}
%\frac{DP}{Dt} 
\frac{\p P}{\p t} + \bm{v}\cdot\nabla P  
&= - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2}
                + \mathcal{C},  \label{eff_energy} \\
\mathcal{C}&\equiv c_s^2 \nabla\cdot\left[\tstop\left(1 -
  \frac{P}{c_s^2\rho}\right)\nabla 
  P\right].
\end{align} 
For comparison with the standard energy equation in hydrodynamics,
re-writing Eq. \ref{tempeq} for a pure ideal gas with $P \propto
\rhog T$, and reverting back to gas velocities, gives  
%For later comparison it is convenient to re-write the temperature
%evolution, Eq. \ref{tempeq}, to that for the pressure: 
\begin{align}
  \frac{\p P}{\p t} + \bm{v}_\mathrm{g}\cdot\nabla P = - \gamma P
  \nabla\cdot \bm{v}_\mathrm{g}  +
  \frac{P}{T}\left(\mathcal{H} -
    \Lambda\right).\label{real_pressure} 
%\frac{\rho \mathcal{R}}{\mu}\frac{T
%  - T_\mathrm{ref}}{t_\mathrm{cool}}, \label{realenergy}
\end{align} 
We can thus interpret Eq. \ref{eff_energy} as the energy equation for
and ideal gas with adiabatic index $\gamma=1$, but now with the
imposed temperature gradient $\nabla c_s^2$ and dust-gas drag $\mathcal{C}$ 
acting as source terms.

%playing the
%role of radiative cooling ($\Lambda$ in 
%Eq. \ref{tempeq}). 
%In this form, the function $\mathcal{C}$ can be interpreted
%as a cooling term.

If we denote 
\begin{align}
  \bm{F} \equiv  - \frac{\nabla P}{\rho}
\end{align}
for the pressure forces, then
\begin{align*}
  \mathcal{C} = - c_s^2\nabla \left( \tepsilon \tstop \rho \bm{F}
  \right), 
\end{align*}
which is in the same form as cooling by radiative diffusion. In protoplanetary
disks the corresponding `heat flux', proportional to $\bm{F}$, is
directly radially outwards and vertically upwards. This is simply a
reflection of particle drift towards increasing pressure (inwards and 
downwards). Particle flux into a region contributes to `cooling' of
that region because the reduced temperature is lowered.

@@

Since we have shown that a locally isothermal dusty gas effectively
has $\gamma=1$ we can define an effective 
entropy for the \emph{mixture} as 
\begin{align}
   \seff \equiv \ln \frac{P}{\rho} = \ln{\left[c_s^2(1-\tepsilon)\right]}.  
\end{align} 
The constant heat capacity has been absorbed into $\seff$. 
%This follows naturally by combining the effective 
%energy Eq. \ref{eff_energy} and Eq. \ref{masseq}. 
Then 
\begin{align*}
  \frac{D \seff}{D t} = \bm{v}\cdot\nabla\ln{c_s^2} +
  \frac{c_s^2}{P}\nabla\cdot\left(\tstop\tepsilon\nabla P\right), 
\end{align*}
which is equivalent to entropy evolution in standard  
hydrodynamics, albeit with source terms. With an effective
entropy defined this way, many of the  
results concerning the stability of (locally) isothermal dusty gas
will have identical form and interpretations to that for pure gas.

@@

%\subsection{Dusty buoyancy forces}
We can now define the vertical buoyancy frequency $N_z$ of the mixture
as    
%$N_z$ of the mixture is given by 
\begin{align}
  N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p \seff}{\p z}
  =c_s^2(r) \frac{\p \ln{\rhog}}{\p z}\frac{\p \tepsilon}{\p z},
%  \quad 
 % N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p \seff}{\p z}.
%&=
  %\frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\frac{\p\ln\rhog}{\p 
%  z}\frac{\p\epsilon}{\p z} \\ &
  %                                =
 % \frac{\epsilon}{\left(1+\epsilon\right)^2}\left(\frac{z}{\Htilde}\right)^2\OmK^2\notag,  
\end{align}
%Typically in a thin, smooth disk we have $|N_r|\ll |N_z|$. 
where the second equality assumes a vertically isothermal disk. 
Protoplanetary disks have $\p_z\rhog<0$, thus stability against
convection ($N_z^2>0$) requires $\p_z\tepsilon <0 $. This is equivalent to
entropy increasing away from the midplane. A similar expression exists
for the radial buoyancy frequency $N_r$, but $|N_r|\ll |N_z|$ in thin,
smooth disks.

@@

%{\bf
\subsection{Physical disk conditions}
%physical disk conditions underwhich the above approx
%(small/well-coupled dust and isothermal gas) is applicable. smaller
%than mm at few tens of au. give dimensional numbers here
%}
The above correspondence between isothermal dusty gas and adiabatic
pure gas is derived under the terminal velocity approximation, which applies
to short stopping times, $\tstop\OmK\ll 1$; and the locally isothermal
approximation, which applies to short cooling times, $t_\mathrm{cool}\OmK\ll
1$. In typical protoplanetary disk
models such as the Minimum Mass Solar Nebula, $\tstop$ decreases for
smaller particles and/or at smaller radii \citep{youdin11}. However,
$t_\mathrm{cool}$ is generally only small in the outer disk
\citep{lin15,malygin17}. Combining these results, we estimate the 
correspondence will be applicable to particles less than mm in size at a few
to 10s of AU in protoplanetary disks. However, note that the
isothermal approximation may be relaxed to other fixed equations of
state (see \S\ref{gen_poly}).

@@

On the other hand, Eq. \ref{dusty_solberg2} can be violated in
disks if the dust is vertically well-mixed ($\p_z\tepsilon =
0$) but  radially-varying ($\p_r\tepsilon\neq0$). In this case the
left-hand-side of Eq. \ref{dusty_solberg2} becomes
\begin{align}
-c_s^4\left(\frac{\p\ln{\rhog}}{\p
    z}\right)^2\left(\frac{\p\tepsilon}{\p r}\right)^2<0.
\end{align}
Such an isothermal dusty disk is \emph{unstable} but there is no
effective vertical buoyancy to stabilize vertical motions ($N_z=0$)
that can tap into the free energy associated with  
vertical shear due to the radial gradient in the dust-to-gas ratio. 
This situation is identical to the pure gas, adiabtic simulations of \cite{nelson13}
where the gas entropy is vertically uniform but has a radial
gradient. The authors indeed find instability. We give a numerical
example in the dusty context in \S\ref{vert_mixed}.

### 2017-08-02 paper.tex
revising according to AY's comments, up to 3.4

The two-fluid equations for dust and gas can be reformulated into 
equivalent dynamical equations for the center of mass velocity $\bm{v}$
and the relative dust-gas velocity  $\Delta\bm{v}\equiv \bm{v}_\mathrm{d}-\bm{v}_\mathrm{g}$ 
\citep{youdin05a}. In this reformulation, continuity equations for dust and gas can be replaced 
by the mass conservation of total density $\rho$ plus the evolution of the dust-to-gas ratio 
(or dust fraction, \citealp{laibe14}).

>>>>

The two-fluid equations for dust and gas can be reformulated into 
equivalent dynamical equations for the center of mass velocity $\bm{v}$
and the relative dust-gas velocity  $\bm{v}_\mathrm{d}-\bm{v}_\mathrm{g}$ 
\citep{youdin05a}. In this reformulation, continuity equations for dust and gas can be replaced 
by the mass conservation of total density $\rho$ plus the evolution of the dust-to-gas ratio 
(or dust fraction, \citealp{laibe14}).

### 2017-08-02 summary.tex
revising according to AY's comments, up to 3.4

The simplified one-fluid formulation, 
Eqs. \ref{masseq}---\ref{tempeq}, can 
produce spuriously-growing modes with frequency $|\sigma|\gtrsim
\Omega$ (see Appendix \S\ref{spurious_epi}). This 
results from neglecting terms of $O(|\widetilde{\bm{v}}|^2)$ after converting
the full two-fluid equations to the one-fluid formulation. The full
one-fluid equations 
%is significantly more complex than our 
%model \cite[e.g.][]{laibe14}. It 
 includes an evolutionary equation for $\widetilde{\bm{v}}$, which has no
 hydrodynamic counter-part \citep{laibe14}. However, it is
 still possible to improve upon 
 the first-order formulation whilst retaining similarilty to  
 standard hydrodynamics.  
%without having the complexity of the full equations

>>>>

The simplified one-fluid formulation, 
Eqs. \ref{masseq}---\ref{tempeq}, can 
produce spuriously-growing modes with frequency $|\sigma|\gtrsim
\Omega$ (see Appendix \S\ref{spurious_epi}). This 
results from neglecting terms of $O(|\bm{v}_\mathrm{d}-\bm{v}_\mathrm{g}|^2)$ after converting
the full two-fluid equations to the one-fluid formulation. The full
one-fluid equations 
%is significantly more complex than our 
%model \cite[e.g.][]{laibe14}. It 
 includes an evolutionary equation for $\widetilde{\bm{v}}\equiv\bm{v}_\mathrm{d}-\bm{v}_\mathrm{g}$, which has no
 hydrodynamic counter-part \citep{laibe14}. However, it is
 still possible to improve upon 
 the first-order formulation whilst retaining similarilty to  
 standard hydrodynamics.  
%without having the complexity of the full equations
